EuPPollNet: A European database of plant-pollinator networks

### 2024-06-06 Manuscript/EuPPollNet_Ms.qmd
edits

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitudes range of studies and bioclimatic regions. Given that this database contains one of the largest sets of plant-pollinator networks collated to date, and considering that the prevalence of nestedness has been evaluated only with a relatively small number of networks [@bascompte2003], we also examine whether plant-pollinator networks are more nested than expected by chance. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. As this database is the result of one of the working packages of the European project Safeguard, first, data was directly asked to members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching studies on Google Scholar of under-represented regions within the database. To keep high quality standards that will allow robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. EXPAND HERE!

Consistent with @olesen2002, we find that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones, as these networks are likely to deviate more from a perfect nested or hierarchical structure compared to networks at higher latitudes. Finally, while one third of networks were more nested than expected by chance, two thirds did not show statistical difference with null expectations. Therefore, while this supports the idea that plant-pollinator networks tend to be nested [@bascompte2003], this result suggests that nestedness could be a less prevalent feature than previously thought for plant-pollinator networks.

>>>>

Here, we present the European Plant-Pollinator Networks database (EuPPollNet), which contains harmonized information on interaction data of plants and pollinators at the European level. The pollinator taxonomic groups are composed by the main orders of entomofauna that visit and pollinate flowering plants in Europe. These include insect species from the orders Hymenoptera, Diptera, Lepidoptera, and Coleoptera [@willemstein1987; @potts2015], which comprise almost the totality of recorded interactions in EuPPollNet (`r format(round(main_orders_percent, 2), nsmall = 2)`%). To understand the scope of the database, we examine the taxonomic and sampling coverage of the different plant and pollinator species at European level with the help of the most up-to-date species checklists and rarefaction analyses. In addition, we investigate how two key structural metrics of plant-pollinator networks, connectance and nestedness, change across the latitudes range of studies and bioclimatic regions. Given that this database contains one of the largest sets of plant-pollinator networks collated to date, and considering that the prevalence of nestedness has been evaluated only with a relatively small number of networks [@bascompte2003], we also examine whether plant-pollinator networks are more nested than expected by chance. EuPPollNet aims to cover a wide range of taxonomic groups and habitats, while also providing other variables of interest that define the ecological context and sampling methods of the study. In addition, EuPPollNet offers a transparent and accessible workflow of its data management and species harmonization that allows the database to be reused and to expand over time. This database provides a large number of community-level networks with curated and harmonized data, distinguishing it from other currently available resources that contain plant-pollinator interactions. We expect that EuPPollNet can be used to evaluate macro-ecological processes in plant-pollinator networks, guide conservation planning, and set a baseline for global change research.

@@

The EuPPollNet database includes published and unpublished studies compiled initially by a wide number of researchers and institutions within the European continent. As this database is the result of one of the working packages of the European project Safeguard, first, data was directly asked from members of the Safeguard project. Second, the request was extended to data owners outside of the project. These other data owners were identified by direct communication with colleagues suggested by Safeguard members and by directly searching for studies on Google Scholar of under-represented regions within the database. To keep high quality standards that will allow robust future ecological research, we only included studies meeting the following criteria: 1) studies that contained time- and geo-referenced records of plant-pollinator interactions; and 2) studies with phyto-centric plant-pollinator networks with quantitative visitation data.

@@

Although Europe contains a much larger number of flowering plants than pollinator species (~4 to 1 ratio), the observed number of pollinator species almost doubled the one of plants in the database. This could be explained by the fact that all networks are phytocentric, resulting in sampling bias towards pollinator species [@jordano2016; @vizentin2018]. While animal-centred sampling is likely to increase the plant:pollinator species ratio [e.g., @encinas2023], the spatial scale and environmental context of the sampled communities will also influence their observed diversity, especially given the ability to move of pollinators and the sessile nature of plants. In addition, we found that the accumulation curve of pollinators per plant species does not saturate, which indicates low redundancy of pollinators and that many are regionally “rare”. Rare pollinators have been proven to be functionally important for plant species at large scales [@winfree2018; @simpson2022], highlighting the need to conduct further sampling events to observe these rare species and to effectively understand and protect plant-pollinator biodiversity.

Consistent with @olesen2002, we find that residual connectance (i.e., the deviation from the expected connectance for a given network size) was lower at higher latitudes, while normalised nestedness, which is negatively correlated with residual connectance, increased towards higher latitudes. Networks at lower latitudes in Europe are exposed to higher temperatures and are bee-dominated, which can result in higher visitation rates [@arroyo1985; @classen2015; @herrera2019], and the overall level of pollinator generalization is known to be higher at lower latitudes [@schleuning2012]. These factors should increase the possible number of connections that can be established between plants and pollinators for a given network size, resulting in more connected networks but less nested ones, as these networks are likely to deviate more from a perfect nested or hierarchical structure compared to networks at higher latitudes. Finally, while one third of networks were more nested than expected by chance, two thirds did not show statistical difference with null expectations. Therefore, while this supports the idea that plant-pollinator networks tend to be nested [@bascompte2003], this result suggests that nestedness could be a less prevalent feature than previously thought for plant-pollinator networks [@payrato2019].
